---
title: "SNP-to-gene mapping"
author: "Fabrício Almeida-Silva"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,
                      warning=FALSE, 
                      message=FALSE,
                      tidy.opts=list(width.cutoff=60),
                      tidy=TRUE)
```

## SNP-to-gene mapping

First, we need to choose which sliding window we will use to pick putative candidate genes. For that, we will perform simulations of sliding windows from 100 kb to 2 Mb with increments of 100 kb.

```{r simulate_windows}
library(ggplot2)
library(here)
library(GenomicRanges)
library(GenomeInfoDb)
library(cageminer)
library(tidyverse)

# Load SNP ranges
load(here("products", "result_files", "snp_granges.rda"))

# Create a gene-based GRanges object
gff <- rtracklayer::import(here("data", "PLAZA_selected.transcripts.gff.gz"))
gff_gene <- gff[gff$type == "gene", ]

# Simulate sliding windows
sims <- simulate_windows(gff_gene, snp_grangeslist)
sims2 <- sims + 
  geom_hline(yintercept=5604, col="darkred", linetype="dashed") +
  annotate("text", x = "1.8", y = 6200, label = "10%", col="darkred")
ggsave(sims2,
       filename = here(
         "products", "plots", "simulation_genes_per_sliding_window.png"
       ), width=9, height=5)
```

Ideally, we should pick all genes in a sliding window of 2 Mb as possible candidates. However, there is trade-off, because the more we increase the number of candidate genes, the more noise we add to our data. Although the number of gene will be high for *Fusarium virguliforme*, we will pick genes in a 2Mb sliding window nevertheless.

```{r get_candidates}
# Extract list of genes in 2 Mb sliding window
candidates <- mine_step1(gff_gene, snp_grangeslist)

# Save object for further analyses
save(candidates, file = here("products", "result_files", "all_candidates.rda"),
     compress="xz")
```

## Candidate gene mining

Now, we will mine candidate genes with {cageminer}. The species we will use are:

- *Phakopsora pachyrhizi*: 19 samples
- *Macrophomina phaseolina*: 4 samples
- *Fusarium virguliforme*: 40 samples
- *Fusarium graminearum*: 11 samples
- *Cadophora gregata*: 27 samples

To start, let's get the guide genes.

```{r get_guides}
mapman <- readr::read_tsv("ftp://ftp.psb.ugent.be/pub/plaza/plaza_public_dicots_03//MapMan/mapman.gma.csv.gz", skip=8)
mapman <- as.data.frame(mapman[, c(2,4)])

#----Convert a1.v1 IDs to a2v1----
ids_v1v2 <- read.csv(here("data", "Gmax_v1_to_v2.txt"), header=TRUE, sep="\t")
annotation_mapman <- merge(mapman, ids_v1v2, by.x="gene_id", by.y="V1")
annotation_mapman <- annotation_mapman[, c(3,2)]
names(annotation_mapman) <- c("Gene", "Description")

#----Get biotic stress-related genes----
guides <- annotation_mapman %>%
  filter(str_detect(
    Description, "ethylene.induced|ethylene.signal|jasmonate.induced|misc.peroxidases|misc.protease inhibitor|protease|phenylpropanoids.lignin|receptor kinases.leucine rich|stress.biotic")
    ) %>%
  mutate(Description = str_replace_all(
    Description,
    c("lignin biosynthesis.*" = "lignin biosynthesis",
      "rich repeat.*" = "rich repeat"))
  ) %>%
  filter(!str_detect(Description, "ubiquitin"))

write_tsv(guides, 
          here("products", "tables", "sup_table3.tsv"))

# EDA: Number of genes per class
guides %>%
  janitor::tabyl(Description) %>%
  arrange(-n)
```

Find high-confidence candidates.

```{r mining}
library(BioNERO)
library(SummarizedExperiment)
set.seed(1)

#----Load and filter expression data----
load("~/Dropbox/Atlas/atlasv2_tpm.rda")
exp <- atlas_tpm[, atlas_tpm$Stress_info == "fungus" &
                   !is.na(atlas_tpm$Stress_info)]
rm(atlas_tpm)

#----Infer a single GCN for all species----
exp_fungi <- exp_preprocess(exp, min_exp=5, Zk_filtering=FALSE)
metadata_fungi <- colData(exp_fungi) %>%
  as.data.frame() %>%
  mutate(Pathogen = str_sub(Pathogen, 1, 3)) %>%
  mutate(annot = paste(Pathogen, Sample_description, sep = "_")) %>%
  select(annot)
colData(exp_fungi) <- DataFrame(metadata_fungi)

sft_fungi <- SFT_fit(exp_fungi, net_type="unsigned", cor_method="pearson")
gcn_fungi <- exp2gcn(exp_fungi, net_type="unsigned", cor_method="pearson",
                     SFTpower=sft_fungi$power)

#----Mine candidates----
# C. gregata
mining_cgregata <- mine_candidates(
  gene_ranges = gff_gene, 
  marker_ranges = snp_grangeslist$Cgregata,
  exp = exp_fungi,
  gcn = gcn_fungi,
  guides = guides,
  metadata = metadata_fungi,
  sample_group = c("Pgr_stress_12h", "Pgr_stress_36h", "Pgr_stress_24h")
)
length(unique(mining_cgregata$gene))

# F. graminearum
mining_fgraminearum <- mine_candidates(
  gene_ranges = gff_gene, 
  marker_ranges = snp_grangeslist$Fgraminearum,
  exp = exp_fungi,
  gcn = gcn_fungi,
  guides = guides,
  metadata = metadata_fungi,
  sample_group = c("Fgr_stress_PI567301B", "Fgr_stress_wyandot")
)
length(unique(mining_fgraminearum$gene))

# F. virguliforme
mining_fvirguliforme <- mine_candidates(
  gene_ranges = gff_gene, 
  marker_ranges = snp_grangeslist$Fvirguliforme,
  exp = exp_fungi,
  gcn = gcn_fungi,
  guides = guides,
  metadata = metadata_fungi,
  sample_group = c("Fvi_stress_0dpi", "Fvi_stress_10-24days",
                   "Fvi_stress_10dpi", "Fvi_stress_14dpi",
                   "Fvi_stress_2dpi", "Fvi_stress_3-5days",
                   "Fvi_stress_4dpi", "Fvi_stress_7dpi")
)
length(unique(mining_fvirguliforme$gene))

# M. phaseolina
mining_mphaseolina <- mine_candidates(
  gene_ranges = gff_gene, 
  marker_ranges = snp_grangeslist$Mphaseolina,
  exp = exp_fungi,
  gcn = gcn_fungi,
  guides = guides,
  metadata = metadata_fungi,
  sample_group = c("Mph_stress_res", "Mph_stress_sus")
)
length(unique(mining_mphaseolina$gene))

# P. pachyrhizi
mining_ppachyrhizi <- mine_candidates(
  gene_ranges = gff_gene, 
  marker_ranges = snp_grangeslist$Ppachyrhizi,
  exp = exp_fungi,
  gcn = gcn_fungi,
  guides = guides,
  metadata = metadata_fungi,
  sample_group = c("Ppa_stress_10dai", "Ppa_stress_14dai", "Ppa_stress_3dai",
                   "Ppa_stress_48hai", "Ppa_stress_7dai", "Ppa_stress_7hai",
                   "Ppa_stress_res_24h", "Ppa_stress_sus_24h")
)
length(unique(mining_ppachyrhizi$gene))

# Save all candidates
save(mining_cgregata, 
     file = here("products", "result_files", "cgregata_minedcand.rda"),
     compress="xz")

save(mining_fgraminearum, 
     file = here("products", "result_files", "fgraminearum_minedcand.rda"),
     compress="xz")

save(mining_fvirguliforme, 
     file = here("products", "result_files", "fvirguliforme_minedcand.rda"),
     compress="xz")

save(mining_mphaseolina, 
     file = here("products", "result_files", "mphaseolina_minedcand.rda"),
     compress="xz")

save(mining_ppachyrhizi, 
     file = here("products", "result_files", "ppachyrhizi_minedcand.rda"),
     compress="xz")

mined_cand_df <- rbind(mining_cgregata, mining_fgraminearum,
                       mining_fvirguliforme, mining_mphaseolina,
                       mining_ppachyrhizi)
mined_cand_df$Species <- c(
  rep("Cgregata", nrow(mining_cgregata)),
  rep("Fgraminearum", nrow(mining_fgraminearum)),
  rep("Fvirguliforme", nrow(mining_fvirguliforme)),
  rep("Mphaseolina", nrow(mining_mphaseolina)),
  rep("Ppachyrhizi", nrow(mining_ppachyrhizi))
)
mined_cand_df <- mined_cand_df[!duplicated(mined_cand_df$gene,
                                           mined_cand_df$Species),
                               c("gene","Species"),drop=FALSE]
write.table(mined_cand_df, 
            file = here("products", "tables", "all_mined_candidates.txt"),
            row.names=FALSE, quote=FALSE)
```

Show candidates in a Venn Diagram.

```{r venn_plot}
venn_pdata <- list(
  Fvirguliforme = unique(mining_fvirguliforme$gene),
  Fgraminearum = unique(mining_fgraminearum$gene),
  Cgregata = unique(mining_cgregata$gene),
  Mphaseolina = unique(mining_mphaseolina$gene),
  Ppachyrhizi = unique(mining_ppachyrhizi$gene)
)

# Plot venn diagram using the VennDiagram package
library(VennDiagram)
venn.diagram(
        x = venn_pdata,
        category.names = c("F. virguliforme" , "F. graminearum" , 
                           "C. gregata", "M. phaseolina", "P. pachyrhizi"),
        filename = here("products", "plots", "venn_diagram_candidates.png"),
        output=TRUE,
        
        # Output features
        imagetype="png" ,
        height = 1600, 
        width = 1600, 
        resolution = 600,
        
        # Circles
        lwd = 2,
        lty = 'blank',
        fill = ggsci::pal_aaas()(5),
        
        # Numbers
        cex = 0.5,
        fontface = "bold",
        fontfamily = "sans",
        
        # Set names
        cat.cex = 0.5,
        cat.fontface = "italic",
        cat.default.pos = "outer",
        cat.fontfamily = "sans",
        margin = 0.2
)
```

For the high-confidence candidates identified, we will create a score for the genes which consists of:
- Multiply correlation by 1.5 if gene is hub OR TF, by 2 if gene is hub AND TF, and by 1 if it is none.

```{r prioritization}
#----Get candidates that are hubs----
hubs <- get_hubs_gcn(exp_fungi, gcn_fungi)
candidates_hubs <- lapply(venn_pdata, function(x) {
  return(hubs[hubs$Gene %in% x, ])
})

#----Get candidates that are TFs----
tfs <- as.data.frame(readr::read_tsv("http://planttfdb.gao-lab.org/download/TF_list/Gma_TF_list.txt.gz"))[, c(2,3)]
candidates_tfs <- lapply(venn_pdata, function(x) {
  y <- tfs[tfs$Gene_ID %in% x, ]
  return(y)
})


#----Score genes----
hubs <- hubs$Gene
tfs <- tfs$Gene_ID

cgregata_score <- score_genes(mining_cgregata, hubs, tfs)
fvirguliforme_score <- score_genes(mining_fvirguliforme, hubs, tfs)
fgraminearum_score <- score_genes(mining_fgraminearum, hubs, tfs)
mphaseolina_score <- score_genes(mining_mphaseolina, hubs, tfs)
ppachyrhizi_score <- score_genes(mining_ppachyrhizi, hubs, tfs)

#----Export best candidates as a table----
dfexport_candidates <- rbind(
  cgregata_score %>% mutate(species = "C. gregata") %>% select(1,5,6),
  fgraminearum_score %>% mutate(species = "F. graminearum") %>% select(1,5,6),
  fvirguliforme_score %>% mutate(species = "F. virguliforme") %>% select(1,5,6),
  mphaseolina_score %>% mutate(species = "M. phaseolina") %>% select(1,5,6),
  ppachyrhizi_score %>% mutate(species = "P. pachyrhizi") %>% select(1,5,6)
)

write_tsv(dfexport_candidates, 
          file = here("products", "tables", "top_candidates.tsv"))
```

To conclude, let's see the descriptions for these genes.

```{r description}
load(here("data", "ath_best_hit.rda"))
go_annotation <- as.data.frame(
  readr::read_tsv("ftp://ftp.psb.ugent.be/pub/plaza/plaza_public_dicots_04_5/GO/go.gma.csv.gz", skip=8)
)
go_annotation <- go_annotation[, c(1,8)]
names(go_annotation) <- c("Gene", "GO_description")

# Get gene description based on A. thaliana descriptions
cand_des <- readr::read_tsv("ftp://ftp.psb.ugent.be/pub/plaza/plaza_public_dicots_04/Descriptions/gene_description.ath.csv.gz", skip=8) %>%
  select(1,3) %>%
  rename(Gene = `#gene_id`, 
         Des = id) %>%
  right_join(ath_best_hit, by = c("Gene" = "Ath_ortho")) %>%
  select(3, 2) %>%
  rename(Gene = Gene.y) %>%
  right_join(dfexport_candidates, by = c("Gene" = "gene")) %>%
  left_join(go_annotation) %>%
  arrange(species)

print(cand_des, n=Inf)
write_tsv(cand_des, 
          file = here("products", "tables", "top_candidates_description.tsv"))
```

Enrichment analysis:

```{r enrichment_analysis_candidates}
# All top genes - GO
sea_topcand <- BioNERO::enrichment_analysis(
  genes = dfexport_candidates$gene,
  background_genes = rownames(exp_fungi),
  annotation = go_annotation
)

# All mined candidate genes - GO
mined_cand <- c(
  venn_pdata$Cgregata, venn_pdata$Fgraminearum,
  venn_pdata$Fvirguliforme, venn_pdata$Mphaseolina,
  venn_pdata$Ppachyrhizi
)

sea_minedcand <- BioNERO::enrichment_analysis(
  genes = mined_cand,
  background_genes = rownames(exp_fungi),
  annotation = go_annotation
)
sea_minedcand[, -6]

# All mined candidates - MapMan
sea_minedcand_mapman <- BioNERO::enrichment_analysis(
  genes = mined_cand,
  background_genes = rownames(exp_fungi),
  annotation = annotation_mapman
)
sea_minedcand_mapman[, -6]

# All mined candidates - Interpro
interpro_annotation <- as.data.frame(
  read_tsv("ftp://ftp.psb.ugent.be/pub/plaza/plaza_public_dicots_04/InterPro/interpro.gma.csv.gz", skip=8)
)
interpro_annotation <- interpro_annotation[, c(1,4)]

sea_minedcand_interpro <- BioNERO::enrichment_analysis(
  genes = mined_cand,
  background_genes = rownames(exp_fungi),
  annotation = interpro_annotation
)
sea_minedcand_interpro[, -6]
```

## Export module enrichment and network file for shiny app development

```{r export_files}
# Create files
mod_enrich <- module_enrichment(
       net = gcn_fungi,
       background_genes = rownames(exp_fungi),
       annotation_mapman)
enrich_go <- module_enrichment(
       net = gcn_fungi,
       background_genes = rownames(exp_fungi),
       go_annotation)
write_tsv(enrich_go,
          file=here::here("products", "result_files", "enrich_go.tsv"))

enrich_interpro <- module_enrichment(
       net = gcn_fungi,
       background_genes = rownames(exp_fungi),
       interpro_annotation)
write_tsv(enrich_interpro,
          file=here::here("products", "result_files", "enrich_interpro.tsv"))

genes_modules <- gcn_fungi$genes_and_modules
degree <- gcn_fungi$kIN

edges <- lapply(unique(genes_modules$Modules), function(x) {
  y <- get_edge_list(gcn_fungi, module = x, filter=TRUE,
                     method="min_cor", rcutoff = 0.4)
  return(y)
})
names(edges) <- unique(genes_modules$Modules)
edges[names(edges) == "grey"] <- NULL

greygenes <- genes_modules[genes_modules$Modules == "grey", 1]
cor_matrix <- gcn_fungi$correlation_matrix[
  !(rownames(gcn_fungi$correlation_matrix) %in% greygenes),
  !(colnames(gcn_fungi$correlation_matrix) %in% greygenes)
]
cor_matrix[cor_matrix < 0.2 & cor_matrix > -0.2] <- NA
edgelist <- cormat_to_edgelist(cor_matrix)

# Export files
save(genes_modules, file=here("products", "result_files", 
                              "shiny_gmodules.rda"), compress="xz")
save(degree, file=here("products", "result_files", 
                              "shiny_degree.rda"), compress="xz")
save(mod_enrich, file=here("products", "result_files", 
                              "shiny_modenrich.rda"), compress="xz")
save(edges, file=here("products", "result_files", 
                              "shiny_edgelists.rda"), compress="xz")
save(hubs, file=here("products", "result_files", 
                              "shiny_hubs.rda"), compress="xz")


```

## Session information

This document was created under the following conditions:

```{r session_info} 
sessionInfo()
```



